A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-04-16 03-semantics.tex
Finishes refactoring Sections 3 and 6

We will refer to the triples on the left-hand and right-hand sides of
symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The description
on the left of~$\trans$ is called the \emph{input description},
the one on its right is called the \emph{output description}.

@@

At the beginning of a reaction to an input event~$id$, the input description is
initialized with stack level~0 ($n=0$) and with the externally emitted event~$e=id$.
%, but \code{emitInt} expressions can increase the stack level.
At the end of a reaction, after an arbitrary but finite number of transitions,
the last output description will block with a possibly modified program~$p'$ at stack
level~0 and no event emitted:
\[
  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
\]

@@

The rules \R{push} and \R{pop} for~$\out$ transitions are non-recursive
definitions that apply to the program as a whole.  These are the only rules
that manipulate the stack level.
\begin{align*}
  &\AxiomC{$e\ne\nil$}
  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
  \DisplayProof
  \Rtag{push}\\[1.2\jot]
  %%
  &\AxiomC{$n>0$}
  \AxiomC{$p=\ceu{\Nop} \vee p=\ceu{\Break} \vee \isblocked(p,n)$}
  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
  \DisplayProof
  \Rtag{pop}
\end{align*}

@@

A complete reaction consists of a series of transitions of the form
\begin{align*}
  \<p,0,e_\ext>\outpush\<p_1,1,\nil>
  \Big[\null\nst[*]\null\out\null\Big]\!\!\ast
  \null\nst[*]\null\outpop\<p',0,\nil>\,,
\end{align*}
%
%-
% \begin{align*}
% a) &\5\5
%     \LL p,0,ext \RR
%         \1\xrightarrow[out]{push}\1
%     \LL q,1,\epsilon \RR
% \\
% b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
%     \LL r,i,e \RR
%         \1\xrightarrow[out]\1
%     \LL s,j,\epsilon \RR \1]*
% \\
% c) &\5\5 \1\xrightarrow[in]{*}\1
%     \LL t,k,\epsilon \RR
%         \1\xrightarrow[out]{pop}\1
%     \LL u,0,\epsilon \RR
% \end{align*}
%-
%
where the~$\ast$'s stand for a finite (possibly zero) number of repetitions.
First, a~$\outpush$ starts a nested reaction at level~1.
Then, a series of alternations between zero or more~$\nst$ transitions (nested reactions) and a
single~$\out$ transition (stack operation) takes place.
Finally, a last~$\outpop$ transition decrements the
stack level to~0 and terminates the reaction.

The~$\nst$ rules for atoms are defined as follows:
\begin{align*}
  \<\ceu{\Mem(\Id)},n,\nil>
  &\nst\<\ceu{\Nop},n,\nil>\Rtag{mem}\\[\JOT]
  %%
  \<\ceu{\EmitInt(\Id)},n,\nil>
  &\nst\<\ceu{\CanRun(n)},n,\ceu{\Id}>\Rtag{emit-int}\\[\JOT]
  %%
  \<\ceu{\CanRun(n)},n,\nil>
  &\nst\<\ceu{\Nop},n,\nil>\Rtag{can-run}
\end{align*}

@@

The rules for conditionals and sequences are the following:
\vskip-.6\baselineskip
\begingroup
\def\JOT{.5\jot}
\begin{gather*}
  \AxiomC{$\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<p,n,\nil>$}
  \DisplayProof
  \Rtag{if-true}\\[\JOT]
  %%
  \AxiomC{$\lnot\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<q,n,\nil>$}
  \DisplayProof
  \Rtag{if-false}\\[\JOT]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\,;\,q},n,\nil>\nst\<\ceu{p';\,q},n,e>$}
  \DisplayProof
  \Rtag{seq-adv}\\[\JOT]
  \<\ceu{\Nop;\,q},n,\nil>\nst\<q,n,\nil>\Rtag{seq-nop}\\[\JOT]
  %%
  \<\ceu{\Break;\,q},n,\nil>\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
\end{gather*}
\endgroup
% \begin{align*}
%   \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[\JOT]
%   %%
%   \<\ceu{\Break;\,q},n,\nil>&\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
% \end{align*}

@@

Transitions~$\out$ and~$\nst$ are defined in such a way that given an input
description either no rule is applicable or exactly one of them can be
applied (no choice involved).  This coupled with the fact that the output of
every rule is a function of its input implies that transitions are
deterministic: the same input description, if it can transition, will always
result in the same output description.  In other words, the transition
relation~$\trans$ is in fact a partial function.

@@

The main result of this section, Theorem~\ref{thm.x.det}, establishes that
any given number~$i\ge0$ of applications of arbitrary transition rules,
starting from the same input description, will always lead to the same
output description, i.e., any finite sequence of transitions behave
deterministically.

\begin{restatable}[Determinism]{theorem}{thmxdet}
  \label{thm.x.det}
  %%
  $\delta\trans[i]\delta_1$ and~$\delta\trans[i]\delta_2$
  implies~$\delta_1=\delta_2$.
\end{restatable}

\begin{proof}
  By induction on~$i$.  The theorem is trivially true if~$i=0$ and follows
  directly from the previous lemmas if~$i=1$.  Suppose
  \[
    \delta\trans[1]\delta_1'\trans[i-1]\delta_1
    \quad\text{and}\quad
    \delta\trans[1]\delta_2'\trans[i-1]\delta_2\,,
  \]
  for some~$i>1$, $\delta_1'$ and~$\delta_2'$.
  %%
  Then, by Lemma~\ref{lem.x.det-out} or~\ref{lem.x.det-nst}, depending on
  whether the first transition is~$\out$ or~$\nst$ (it cannot be both),
  $\delta_1'=\delta_2'$, and by the induction hypothesis,
  $\delta_1=\delta_2$.
\end{proof}

@@

We now turn to the problem of termination.  We want to show that any
sufficiently long sequence of applications of arbitrary transition rules
will eventually lead to an irreducible description, i.e., one that cannot be
modified by further transitions.  Before doing that, however, we need to
introduce some notation and establish some basic properties of
transitions~$\nst$ and~$\out$.

@@

Theorem~\ref{thm.x.term-nst-*} establishes that a finite (possibly zero)
number of~$\nst$ transitions eventually leads to a nested-irreducible
description.  Put another way, it establishes that, for any input
description~$\delta$, it is always possible to transform~$\delta$ in a
nested-irreducible description~$\delta'$ by applying to it a sufficiently
long sequence of~$\nst$ transitions.  The proof follows by induction on the
structure of programs (members of set~$P$) and depends on
Lemma~\ref{lem.x.props-nst-i} and Assumption~\ref{ass.x.syn-rest}.

@@

The main result of this section, Theorem~\ref{thm.x.term}, is similar to
Theorem~\ref{thm.x.term-nst-*} but applies to transitions~$\trans$ in
general.  Before stating and proving it, we need to characterize irreducible
descriptions in general.  This characterization, Definition~\ref{def.x.H},
and depends on the notions of potency and rank.

@@

  The function~$\pot'$ is defined by the following clauses:
  \begin{enumerate:a}
  \item\label{def.x.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$.
    %%
  \item$\pot'(\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}})=$\\
    \strut\quad$\max\{\pot'(p_1),\pot'(p_2)\}$.
    %%
  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$.
    %%
  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(p_1)+\pot'(p_2)$.
    %%
  \item$\pot'(\ceu{{p_1}\Or{p_2}})=\pot'(p_1)+\pot'(p_2)$.
    %%
  \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
    \begin{align*}
      \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\
      \pot'(\ceu{p_1\AtLoop{p_2}})&=
      \begin{cases}
        \pot'(p_1)              &\text{if\enspace(\dag)}\\
        \pot'(p_1)+\pot'(p_2)   &\text{otherwise,}\\
      \end{cases}
    \end{align*}
    where~(\dag) stands for: ``a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ occurs
    in all execution paths of~$p_1$''.
    %%
  \item If~$p_1,p_2\ne\ceu{\Break}$,
    \[
      \pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
    \]
    %%
  \item\label{def.x.pot.before-last} If~$p_1,p_2\ne\ceu{\Break}$
    and~$p_1,p_2\ne\ceu{\Nop}$,
    \[
      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
    \]
    %%
  \item\label{def.x.pot.last} Otherwise, if none
    of~\ref{def.x.pot.first}--\ref{def.x.pot.before-last} applies,
    $\pot(\_)=0$.
  \end{enumerate:a}
\end{definition}

\begin{definition}
  \label{def.x.rank}
  %%
  The \emph{rank} of a description~$\delta=\<p,n,e>$,
  denoted~$\rank(\delta)$, is a pair of nonnegative integers~$\<i,j>$ such
  that
  \begin{alignat*}{2}
    i&=\pot(p,e) &\quad\text{and}\quad
    j&=
       \begin{cases}
         n  &\text{if~$e=\nil$}\\
         n+1&\text{otherwise\,.}
       \end{cases}
  \end{alignat*}
\end{definition}

Intuitively, the rank of a description~$\delta$ is a measure of the maximum
amount of ``work'' (transitions) required to transform~$\delta$ into an
irreducible description, as defined bellow.

@@

The next two lemmas establish that a single application of~$\out$ or~$\nst$
either preserves or decreases the rank of the input description.  All rank
comparisons assume lexicographic order, i.e., if~$\rank(\delta)=\<i,j>$
and~$\rank(\delta')=\<i',j'>$ then~$\rank(\delta)>\rank(\delta')$ iff~$i>i'$
or~$i=i$ and~$j>j'$.
%%
The proof of Lemma~\ref{lem.x.rank-out} follows directly from~\R{push}
and~\R{pop} and from Definitions~\ref{def.x.pot} and~\ref{def.x.rank}.  The
proof of Lemma~\ref{lem.x.rank-nst}, however, is by induction on the
structure of~$\nst$ derivations.  Since Lemma~\ref{lem.x.rank-nst} is of
fundamental importance for the termination result, we present a sketch of its
proof.

\begin{lemma}\strut
  \label{lem.x.rank-out}
  %%
  \begin{enumerate:a}
  \item\label{lem.rank-out-push} If~$\delta\outpush\delta'$
    then~$\rank(\delta)=\rank(\delta')$.
  \item\label{lem.rank-out-pop} If~$\delta\outpop\delta'$
    then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate:a}
\end{lemma}

\begin{lemma}
  \label{lem.x.rank-nst}
  %%
  If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{lemma}
\begin{proof}
  By induction on the structure of~$\nst$ derivations.  Let
  \[
    \delta=\<p,n,e>
    \quad\text{and}\quad
    \delta'=\<p',n',e'>
  \]
  be descriptions with~$\rank(\delta)=\<i,j>$ and~$\rank(\delta')=\<i',j'>$
  such that~$\delta\nst\delta'$.  Then there is a derivation~$\pi$ whose
  conclusion is~$\delta\nst\delta'$, i.e.,
  $\pi\Vdash\<p,n,e>\nst\<p',n',e'>$.
  %%
  % \[
  %   \pi\Vdash\<p,n,e>\nst\<p',n',e'>\,.
  % \]
  %%
  By definition of~$\nst$, $e=\nil$ and~$n=n'$.  Depending on the structure
  of program~$p$, there are~11 distinct cases to be considered.  In each one
  of them, we need to show that~$\rank(\delta)\ge\rank(\delta')$.

  We proceed to prove the case where~$p=\ceu{{p_1}\AtLoop{p_2}}$, which
  itself consists of three subcases.
  \begin{case}
    $p_1=\ceu{\Nop}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-nop}.  i.e., its conclusion is
    obtained by an application of this rule.  Hence~$p'=p_2$ and~$e'=\nil$.
    Thus
    \begin{align*}
        \rank(\delta)=\<\pot'(p_1)+\pot'(p_2),n>
                     \ge\<pot'(p_2),n>=\rank(\delta')\,.
    \end{align*}
  \end{case}
  \begin{case}
    $p_1=\ceu{\Break}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-brk}.
    Hence~$p'=p_1$ and~$e'=\nil$.  Thus
    $\rank(\delta)=\rank(\delta')=\<0,n>$.
    %%
    % \[
    %   \rank(\delta)=\rank(\delta')=\<0,n>\,.
    % \]
    %%
  \end{case}
  \begin{case}
    $p_1\ne\ceu{\Nop},\ceu{\Break}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-adv}.  Hence there is a
    derivation~$\pi'$ such that
    \begin{equation}
      \label{lem.x.rank-nst-eq1}
      \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
    \end{equation}
    for some~$p_1'$ and~$e_1'$.  Thus~$p'=\ceu{p_1'\AtLoop{\,p_2}}$
    and~$e'=e_1'$.

    There are two subcases.
    \begin{subcase}
      \label{lem.x.rank-nst-case1}
      $\pot'(p)=\pot'(p_1)$.
      %%
      Then, by Definition~\ref{def.pot}, every execution path of~$p_1$
      contains a (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
      A single~$\nst$ cannot terminate the loop, since~$p_1\ne\ceu{\Break}$,
      nor it can consume an~$\ceu{\AwaitExt}$, which means that all
      execution paths in~$p_1'$ still contain a~$\ceu{\Break}$
      or~$\ceu{\AwaitExt}$.  Hence~$\pot'(p')=\pot'(p_1')$.

      There are two subcases.
      \begin{subsubcase}
        $e'=\nil$.
        %%
        Then $\rank(\delta)=\<pot'(p_1),n>$ and
        $\rank(\delta')=\<pot'(p_1'),n>$.  From~\eqref{lem.x.rank-nst-eq1},
        by the induction hypothesis, $\pot'(p_1)\ge\pot(p_1')$.  Thus
        $\rank(\delta)\ge\rank(\delta')$.
        %%
        % \[
        %   \rank(\delta)\ge\rank(\delta')\,.
        % \]
        %%
      \end{subsubcase}
      \begin{subsubcase}
        $e'\ne\nil$.
        %%
        Then~$\pi'$ contains one application of~\R{emit-int} which consumes
        one~$\ceu{\EmitInt(e')}$ expression from~$p_1$ and which implies
        $\pot'(p_1)>\pot'(p_1')$.  Thus
        \[
          \hskip1em\rank(\delta)\!=\!\<\pot'(p_1),n>
          >\<\pot'(p_1'),n+1>\!=\!\rank(\delta')\,.
        \]
        %%
        % \begin{align*}
        %   \rank(\delta)&=\<\pot'(p_1),n>\\
        %                &>\<\pot'(p_1'),n+1>=\rank(\delta')\,.
        % \end{align*}
        %%
      \end{subsubcase}
    \end{subcase}
    \begin{subsubcase}
      $\pot'(p)=\pot'(p_1)+\pot'(p_2)$.
      %%
      Then some execution path in~$p_1$ does not contain a
      (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
      Since~$p_1\ne\ceu{\Nop}$, a single~$\nst$ cannot restart the loop,
      which means that~$p_1'$ still contain some execution path in which
      a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ does not occur.
      Hence~$\pot'(p')=\pot'(p_1')+\pot'(p_2)$.  The rest of this proof is
      similar to that of Case~\ref{lem.x.rank-nst-case1}.\qedhere
    \end{subsubcase}
  \end{case}
\end{proof}

The next theorem is a generalization of Lemma~\ref{lem.x.rank-nst}
for~$\nst[*]$ transitions.  Its proof is by induction on~$i$.

\begin{theorem}
  \label{thm.x.rank-nst-*}
  %%
  If~$\delta\nst[*]\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{theorem}

>>>>

We refer to the triples on the left-hand and right-hand sides of
symbol~$\trans$ as \emph{descriptions} (denoted~$\delta$).  The description
on the left of~$\trans$ is called the \emph{input description}, the one on
its right is called the \emph{output description}.

@@

At the beginning of a reaction to an input event~$id$, the input description
is initialized with stack level~0 ($n=0$) and with the externally emitted
event~$e=\ceu{\Id}$.  At the end of a reaction, after an arbitrary but
finite number of transitions, the last output description will block with a
possibly modified program~$p'$ at stack level~0 and no event
emitted\footnote{We write~$\trans[i]$ to mean~$i$ transitions in sequence,
  and we write~$\trans[*]$ to mean a finite (possibly zero) number of
  transitions in sequence.}:
\[
  \<p,0,e>\mathbin{\trans[*]}\<p',0,\nil>\,.
\]

@@

The rules \R{push} and \R{pop} for~$\out$ transitions are non-recursive
definitions that apply to the program as a whole.  These are the only rules
that manipulate the stack level.
\begin{gather*}
  \AxiomC{$e\ne\nil$}
  \UnaryInfC{$\<p,n,e>\out\<\bcast(p,e),n+1,\nil>$}
  \DisplayProof
  \Rtag{push}\\[1.2\jot]
  %%
  \AxiomC{$n>0$}
  \AxiomC{$p=\ceu{\Nop} \vee p=\ceu{\Break} \vee \isblocked(p,n)$}
  \BinaryInfC{$\<p,n,\nil>\out\<p,n-1,\nil>$}
  \DisplayProof
  \Rtag{pop}
\end{gather*}

@@

A complete reaction consists of a series of transitions% of the form
\begin{align*}
  \<p,0,e_\ext>\outpush\<p_1,1,\nil>
  \Big[\null\nst[*]\null\out\null\Big]\!\!\ast
  \null\nst[*]\null\outpop\<p',0,\nil>\,,
\end{align*}
%
%-
% \begin{align*}
% a) &\5\5
%     \LL p,0,ext \RR
%         \1\xrightarrow[out]{push}\1
%     \LL q,1,\epsilon \RR
% \\
% b) &\5\5 \1[ \1\xrightarrow[in]{*}\1
%     \LL r,i,e \RR
%         \1\xrightarrow[out]\1
%     \LL s,j,\epsilon \RR \1]*
% \\
% c) &\5\5 \1\xrightarrow[in]{*}\1
%     \LL t,k,\epsilon \RR
%         \1\xrightarrow[out]{pop}\1
%     \LL u,0,\epsilon \RR
% \end{align*}
%-
%
First, a~$\outpush$ starts a nested reaction at level~1.  Then, a series of
alternations between zero or more~$\nst$ transitions (nested reactions) and
a single~$\out$ transition (stack operation) takes place.  Finally, a
last~$\outpop$ transition decrements the stack level to~0 and terminates the
reaction.

The~$\nst$ rules for atoms are defined as follows:
\begingroup
\def\JOT{-.1\jot}
\begin{align*}
  \<\ceu{\Mem(\Id)},n,\nil>
  &\nst\<\ceu{\Nop},n,\nil>\Rtag{mem}\\[\JOT]
  %%
  \<\ceu{\EmitInt(\Id)},n,\nil>
  &\nst\<\ceu{\CanRun(n)},n,\ceu{\Id}>\Rtag{emit-int}\\[\JOT]
  %%
  \<\ceu{\CanRun(n)},n,\nil>
  &\nst\<\ceu{\Nop},n,\nil>\Rtag{can-run}
\end{align*}
\endgroup

@@

The rules for conditionals and sequences are the following:
\vskip-.6\baselineskip
\begingroup
\def\JOT{-.1\jot}
\begin{gather*}
  \AxiomC{$\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<p,n,\nil>$}
  \DisplayProof
  \Rtag{if-true}\\[\JOT]
  %%
  \AxiomC{$\lnot\eval(\ceu{\Mem(\Id)})$}
  \UnaryInfC{$\<\ceu{\IfElse{\Mem(\Id)}{p}{q}},n,\nil>\nst\<q,n,\nil>$}
  \DisplayProof
  \Rtag{if-false}\\[\JOT]
  %%
  \AxiomC{$\<p,n,\nil>\nst\<p',n,e>$}
  \UnaryInfC{$\<\ceu{p\,;\,q},n,\nil>\nst\<\ceu{p';\,q},n,e>$}
  \DisplayProof
  \Rtag{seq-adv}\\[\JOT]
  \<\ceu{\Nop;\,q},n,\nil>\nst\<q,n,\nil>\Rtag{seq-nop}\\[\JOT]
  %%
  \<\ceu{\Break;\,q},n,\nil>\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
\end{gather*}
\endgroup
% \begin{align*}
%   \<\ceu{\Nop;\,q},n,\nil>&\nst\<q,n,\nil>\Rtag{seq-nop}\\[\JOT]
%   %%
%   \<\ceu{\Break;\,q},n,\nil>&\nst\<\ceu{\Break},n,\nil>\Rtag{seq-brk}
% \end{align*}

@@

Transitions~$\out$ and~$\nst$ are defined in such a way that given an input
description either no rule is applicable or exactly one of them can be
applied (no choice involved).  This coupled with the fact that the output of
every rule is a function of its input implies that transitions are
deterministic: the same input description, if it can transition, will always
result in the same output description.  Thus the transition
relation~$\trans$ is in fact a partial function.

@@

The main result of this section, Theorem~\ref{thm.x.det}, establishes that
any given number~$i\ge0$ of applications of arbitrary transition rules,
starting from the same input description, will always lead to the same
output description.  In other words, any finite sequence of transitions
behave deterministically.

\begin{restatable}[Determinism]{theorem}{thmxdet}
  \label{thm.x.det}
  %%
  $\delta\trans[i]\delta_1$ and~$\delta\trans[i]\delta_2$
  implies~$\delta_1=\delta_2$.
\end{restatable}
\begin{proof}
  By induction on~$i$.  The theorem is trivially true if~$i=0$ and follows
  directly from the previous lemmas if~$i=1$.  Suppose
  \[
    \delta\trans[1]\delta_1'\trans[i-1]\delta_1
    \quad\text{and}\quad
    \delta\trans[1]\delta_2'\trans[i-1]\delta_2\,,
  \]
  for some~$i>1$, $\delta_1'$ and~$\delta_2'$.
  %%
  Then, by Lemma~\ref{lem.x.det-out} or~\ref{lem.x.det-nst}, depending on
  whether the first transition is~$\out$ or~$\nst$ (it cannot be both),
  $\delta_1'=\delta_2'$, and by the induction hypothesis,
  $\delta_1=\delta_2$.
\end{proof}

@@

We now turn to the problem of termination.  We want to show that any
sufficiently long sequence of applications of arbitrary transition rules
will eventually lead to an irreducible description, i.e., one that cannot be
modified by further transitions.  Before doing that, however, we need to
introduce some notation and establish some basic properties of
the transition relations~$\nst$ and~$\out$.

@@

Theorem~\ref{thm.x.term-nst-*} establishes that a finite (possibly zero)
number of~$\nst$ transitions eventually leads to a nested-irreducible
description.  Hence, for any input description~$\delta$, it is always
possible to transform~$\delta$ in a nested-irreducible description~$\delta'$
by applying to it a sufficiently long sequence of~$\nst$ transitions.  The
proof of the theorem follows by induction on the structure of programs
(members of set~$P$) and depends on Lemma~\ref{lem.x.props-nst-i} and
Assumption~\ref{ass.x.syn-rest}.

@@

The main result of this section, Theorem~\ref{thm.x.term}, is similar to
Theorem~\ref{thm.x.term-nst-*} but applies to transitions~$\trans$ in
general.  Before stating and proving it, we need to characterize irreducible
descriptions in general.  This characterization, given in
Definition~\ref{def.x.H}, depends on the notions of potency and rank.

@@

  Function~$\pot'$ is defined by the following clauses:
  \begin{enumerate:a}
  \item\label{def.x.pot.first}$\pot'(\ceu{\EmitInt}(e))=1$.
    %%
  \item$\pot'(\ceu{\IfElse{\Mem(\Id)\!}{p_1\!}{p_2\!}})\!=\!
    \max\{\pot'(p_1),\pot'(p_2)\}$.
    %%
  \item$\pot'(\ceu{\Loop{p_1}})=\pot'(p_1)$.
    %%
  \item$\pot'(\ceu{{p_1}\And{p_2}})=\pot'(\ceu{{p_1}\Or{p_2}})=
    \pot'(p_1)+\pot'(p_2)$.
    %%
  \item If~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
    \begin{align*}
      \pot'(\ceu{p_1;\,p_2})&=\pot'(p_1)+\pot'(p_2)\\[-.5\jot]
      \pot'(\ceu{p_1\AtLoop{p_2}})&=
      \begin{cases}
        \pot'(p_1)              &\text{if\enspace(\dag)}\\[-.5\jot]
        \pot'(p_1)+\pot'(p_2)   &\text{otherwise,}\\[-.5\jot]
      \end{cases}
    \end{align*}
    where~(\dag) stands for: ``a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ occurs
    in all execution paths of~$p_1$''.
    %%
  \item If~$p_1,p_2\ne\ceu{\Break}$,
    $\pot'(\ceu{{p_1}\AtAnd{p_2}})=\pot'(p_1)+\pot'(p_2)$.
    %%
  \item\label{def.x.pot.before-last} If~$p_1,p_2\ne\ceu{\Break}$
    and~$p_1,p_2\ne\ceu{\Nop}$,
    \vskip-.95em
    \[
      \pot'(\ceu{{p_1}\AtOr{p_2}})=\pot'(p_1)+\pot'(p_2)\,.
    \]
    %%
  \item\label{def.x.pot.last} Otherwise, if none
    of~\ref{def.x.pot.first}--\ref{def.x.pot.before-last} applies,
    $\pot'(\_)=0$.
  \end{enumerate:a}
\end{definition}

\begin{definition}
  \label{def.x.rank}
  %%
  The \emph{rank} of a description~$\delta=\<p,n,e>$,
  denoted~$\rank(\delta)$, is a pair of nonnegative integers~$\<i,j>$ such
  that
  \begin{alignat*}{2}
    i&=\pot(p,e) &\quad\text{and}\quad
    j&=
       \begin{cases}
         n  &\text{if~$e=\nil$}\\
         n+1&\text{otherwise\,.}
       \end{cases}
  \end{alignat*}
Intuitively, the rank of a description~$\delta$ is a measure of the maximum
amount of ``work'' (transitions) required to transform~$\delta$ into an
irreducible description, in the following sense.
\end{definition}

@@

The next two lemmas establish that a single application of~$\out$ or~$\nst$
either preserves or decreases the rank of the input description.  All rank
comparisons assume lexicographic order, i.e., if~$\rank(\delta)=\<i,j>$
and~$\rank(\delta')=\<i',j'>$ then~$\rank(\delta)>\rank(\delta')$ iff~$i>i'$
or~$i=i$ and~$j>j'$.
%%
The proof of Lemma~\ref{lem.x.rank-out} follows directly from~\R{push}
and~\R{pop} and from Definitions~\ref{def.x.pot} and~\ref{def.x.rank}.  The
proof of Lemma~\ref{lem.x.rank-nst}, however, is by induction on the
structure of~$\nst$ derivations.

\begin{restatable}{lemma}{lemxrankout}\strut
  \label{lem.x.rank-out}
  %%
  \begin{enumerate:a}
  \item\label{lem.x.rank-out-push} If~$\delta\outpush\delta'$
    then~$\rank(\delta)=\rank(\delta')$.
  \item\label{lem.x.rank-out-pop} If~$\delta\outpop\delta'$
    then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate:a}
\end{restatable}

%\vskip-.25em

\begin{restatable}{lemma}{lemxranknst}
  \label{lem.x.rank-nst}
  %%
  If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{restatable}

The next theorem is a generalization of Lemma~\ref{lem.x.rank-nst}
for~$\nst[*]$.  Its proof follows from the lemma by induction on~$i$.

\begin{restatable}{theorem}{thmxranknstx}
  \label{thm.x.rank-nst-*}
  %%
  If~$\delta\nst[*]\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{restatable}

We now state and prove the main result of this section,
Theorem~\ref{thm.x.term}, the termination theorem for~$\trans[*]$.  The idea
of the proof is that a sufficiently large sequence of~$\nst$ and~$\out$
transitions eventually decreases the rank of the current description until
an irreducible description is reached.  This irreducible description is the
final result of the reaction.
%%
\begin{restatable}[Termination]{theorem}{thmxterm}
  \label{thm.x.term}
  %%
  For any~$\delta$, there is a~$\delta'_\#$ such
  that~$\delta\trans[*]\delta'_\#$.
\end{restatable}
\begin{proof}
  By lexicographic induction on~$\rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
  $\rank(\delta)=\<i,j>$.

  For the basis, suppose $\<i,j>=\<0,0>$.  Then~$\delta$ cannot be advanced
  by~$\out$, as~$j=0$ implies~$e=\nil$ and~$n=0$.  If~$\delta$ is
  nested-irreducible, the theorem is trivially true,
  as~$\delta\nst[0]\delta_\Hnst$ and~$\delta_\#$.  If~$\delta$ is not
  nested-irreducible then, by Theorem~\ref{thm.x.term-nst-*},
  $\delta\nst[*]\delta'_\Hnst$, for some~$\delta'_\Hnst$.  By
  Theorem~\ref{thm.x.rank-nst-*}, $\rank(\delta)\ge\rank(\delta')$, which
  implies~$\rank(\delta')=\<0,0>$, and so~$\delta'_\#$.

  For the inductive step, suppose~$\<i,j>>\<0,0>$.
  %%
  Then, depending on whether or not $\delta$~is nested-irreducible, there
  are two cases.
  \begin{casex}
  \item\label{thm.x.term.Hnst}$\delta$~is nested-irreducible.
    %%
    If~$j=0$, by Definition~\ref{def.x.H}, $\delta_\#$, and
    so~$\delta\trans[0]\delta_\#$.  If~$j>0$, there are two subcases:
    \begin{casex}
    \item\label{thm.x.term.Hnst-j>0-nonnil}$e\ne\nil$.
      %%
      Then, by~\R{push} and by Theorem~\ref{thm.x.term-nst-*}, there
      are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
      $\delta\outpush\delta_1'\nst[*]\delta'_\Hnst$.  Thus, by
      Lemma~\ref{lem.x.rank-out} and by Theorem~\ref{thm.x.rank-nst-*},
      \begin{align*}
        \rank(\delta)=\rank(\delta_1')=\<i,j>
        \ge\rank(\delta')=\<i',j'>\,.
      \end{align*}
      If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof is
      similar to that of Case~\ref{thm.x.term.Hnst-j>0-nil} below.
      Otherwise, if~$e'\ne\nil$, then $i>i'$, as an~$\ceu{\EmitInt(e')}$ was
      consumed by the nested transitions.  Thus,
      $\rank(\delta)>\rank(\delta')$.  By the induction hypothesis,
      $\delta'\trans[*]\delta''_\#$, for some~$\delta''_\#$.  Therefore,
      $\delta\trans[*]\delta''_\#$.
      %%
    \item\label{thm.x.term.Hnst-j>0-nil}$e=\nil$.
      %%
      Then, as~$j>0$, $\delta\outpop\delta'$, for some~$\delta'$.  By
      Lemma~\ref{lem.x.rank-out}, $\rank(\delta)>\rank(\delta')$.  Hence, by
      the induction hypothesis, $\delta'\trans[*]\delta''_\#$, for
      some~$\delta''_\#$.  And so, $\delta\trans[*]\delta''_\#$.
    \end{casex}

  \item$\delta$~is not nested-irreducible.
    %%
    Then~$e=\nil$ and, by Theorems~\ref{thm.x.term-nst-*}
    and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such
    that~$\delta\nst[*]\delta'_\Hnst$
    with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof is
    similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
  \end{casex}
\end{proof}
